---
title: "Workshop"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workshop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installing required packages

*must have: RStudio with up to date R*

1.  Install the "pacman" package which will facilitate the installation of all other required packages:

    ```{r}
    #install.packages("pacman")
    ```

2.  Install all the required packages using pacman::p_load() - it should skip any packages that you already have installed.

    ```{r}
    pacman::p_load(dplyr,forcats,gdistance,ggspatial,magrittr,memoise,methods,
                      nleqslv,purrr,raster,readr,readxl,rlang,sf,sp,stats,stringr,
                      tabularaster,tibble,tidyr,utils,ggplot2,gridExtra,sysfonts)
    ```

# How to install seabORD

1.  unzip folder on computer and know the path to it (e.g.,):

    "C:/Users/madtig/Desktop/seabORD_pkg-main"

2.  open R Studio (not in the project of the zip file necessarily; just opening R should be enough)

    Run the following code in the console with your defined path:

```{r}
path <- file.path("C:","Users","madtig","Desktop","seabORD_pkg-main", fsep = .Platform$file.sep)

#install.packages(pkgs = path, repos = NULL, type = "source")
```

4.  load the package in an R script:

```{r}
library("seabORD")
```

5.  you can now follow the examples!
