---
title: "Example kittiwake:  step 4 - interpreting the outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example kittiwake:  step 4 - interpreting the outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load the seabORD package

```{r setup, warning = FALSE, message = FALSE}
library(seabORD)
library(ggplot2)
library(magrittr) 
```

```{r}

#use coastline
uk_map <- seabORD::cef_coast_4326


### theme definition
sysfonts::font_add_google("Montserrat", "montserrat")  # Base and heading font
sysfonts::font_add_google("JetBrains Mono", "jetbrains_mono")  # Code font
showtext::showtext_auto()

theme_bslib <- ggplot2::theme(
  plot.background = ggplot2::element_rect(fill = "#fff", color = NA),       # Background color
  panel.background = ggplot2::element_rect(fill = "#EAEFEC", color = NA),  # Panel background (secondary)
  panel.grid.major = ggplot2::element_line(color = "#EAEFEC"),             # Major grid lines
  panel.grid.minor = ggplot2::element_line(color = "#EAEFEC"),             # Minor grid lines
  axis.text = ggplot2::element_text(color = "#292C2F", family = "montserrat"),  # Axis text (foreground and font)
  axis.title = ggplot2::element_text(color = "#292C2F", family = "montserrat"), # Axis titles
  plot.title = ggplot2::element_text(color = "#0483A4", family = "montserrat", size = 16, face = "bold", hjust = 0.5), # Title
  legend.background = ggplot2::element_rect(fill = "#fff", color = NA),    # Legend background
  legend.key = ggplot2::element_rect(fill = "#EAEFEC", color = NA),        # Legend key
  legend.text = ggplot2::element_text(color = "#292C2F", family = "montserrat"), # Legend text
  plot.caption = ggplot2::element_text(color = "#292C2F", family = "jetbrains_mono"),
  axis.title.x = ggplot2::element_blank(),  # Remove x-axis label
  axis.title.y = ggplot2::element_blank(),# Caption for code font
)
```

# Output - reading and interpretation

The main output of seabORD is a list, where the different elements relate to respective parts of the model e.g., wind farm effects, adult outputs, chick outputs etc.

We shall start by reading in an example created from a similar (but not identical) run with the same species (kittiwake), colony (Isle of May), and wind farm set up (Neart na Gaoithe and Inch Cape).

## Read in example output

```{r}
sbordout <- seabORD::example_scenario_output

# str(sbordout) # check out the structure
# names(sbordout)
```

As you can see this is an extensive output with 16 elements to the list. The first five elements of the list are related to inputs ("switches", "Parameters", "ordPar", "modPar", "thisRun") and we include this information so that the runs will be reproducible and are transparent for anyone inspecting the results.

The following 11 outputs, aside from "BirdFlightMap" are all outputs relating to different components/entities in the model and each element typically has a \$data section where the actual data is contained and a \$metadata section which describes each of the columns in the output.

-   "output_f0" - Flight metrics relating to ORD interaction

-   "output_c0" - Chick metrics e.g., number alive and mean body mass

-   "output_a0" - Adult metrics e.g., survival and body mass

-   "output_y0" - Survival metrics relating to calc_pSurvival function which estimates over-winter survival from mass-survival relationships

-   "BirdFlightMap" - a flight map of locations - note this is just for visual assessment

-   "output_i0"

-   "output_i1"

-   "output_i2"

-   "output_i3"

-   "output_i4"

-   "output_i5"

## Plot BirdFlightMap

Let's plot the flight map location output just to see that the run has performed as intended

```{r}
# View the birdfligthmap output which is saved to it:
BirdFlightMap <- 
    terra::rast(sbordout$BirdFlightMap) %>%
  terra::project(terra::crs(uk_map))
BirdFlightMap_df <- as.data.frame(BirdFlightMap, xy = TRUE)

ggplot2::ggplot() + 
  ggplot2:: geom_raster(data = BirdFlightMap_df, ggplot2::aes(x = x, y = y, fill = layer)) +
  # Add the reprojected UK boundary
  ggplot2::geom_sf(data = uk_map, fill = NA, color = "#D63333", size = 1.5) +
  # Customize the plot
  theme_bslib +  #palette theme
    ggplot2::scale_fill_viridis_c(
    na.value = "gray",  # Set the color for NA (i.e., for 0 values)
    name = "BirdFlightMap", direction = -1) +
  ggplot2::labs(fill = "BirdFlightMap locations",
               title = "coasts (red) and BirdFlightMap")+
  # Crop the map to the Firth of Forth and a bit of the North Sea
  ggplot2::coord_sf(
    xlim = c(-3.5, 0),  # Longitude range (adjust as needed)
    ylim = c(55.25, 57.25)     # Latitude range (adjust as needed)
  )
  
```

From this plot we can see clearly the two ORD footprints, with a higher density of foraging location in the 5km displacement zone surrounding each ORD. Some birds have foraging inside the ORD footprints - these will have been indiviudals which were not assigned as displacement susceptible.

## Adult output (output_a0)

```{r}
# inspect adult metric descriptions 
sbordout$output_a0$metadata

# Isolate the adult output:
adults <- sbordout$output_a0$data

# Inspect the dataframe in a new page:
#View(adults)

```

You will see that there are 4 replicates of prey values here with respective rows for baseline ("Season\$base") and scenario ("Season\$scen") matched runs, giving a total of 8 rows.

Conceivably, the average condition of adults across scenario runs at the end of the simulation (BM_adult.mn), will be lower than that for baseline runs. So let's plot that and have a look...

```{r}
ggplot2::ggplot()+
  ggplot2::geom_boxplot(data=adults, aes(x=Season, y=BM_adult.mn)) +
  ggplot2::theme_classic()
```

## Chick output (output_c0)

```{r}
# inspect adult metric descriptions 
sbordout$output_c0$metadata

# Isolate the adult output:
chicks <- sbordout$output_c0$data

# Inspect the dataframe in a new page:
#View(chicks)

```

Conceivably, the number of survivy chicks across scenario runs at the end of the simulation (N_alive_ch), will be lower than that for baseline runs...

```{r}
ggplot2::ggplot()+
  ggplot2::geom_boxplot(data=chicks, aes(x=Season, y=N_alive_ch)) +
  ggplot2::theme_classic()
```

## Flight related output (output_f0)

```{r}
# inspect adult metric descriptions 
sbordout$output_f0$metadata

# Isolate the adult output:
flights <- sbordout$output_f0$data

# Inspect the dataframe in a new page:
#View(flights)

```

"Tot_basickm.mn" is the mean per replicate of how many km birds flew - which would conceivably be higher for scenario runs due to barrier effects increasing flight distance, so let's do another basic plot:

```{r}
ggplot2::ggplot()+
  ggplot2::geom_boxplot(data=flights, aes(x=Season, y=Tot_basickm.mn)) +
  ggplot2::theme_classic()

# and we can see that adults in the scenario run typically flew around 50km more on average over the simulation duration than birds with no wind farm interaction.
```

## Outputs relating to impacts (i0...5)

-   Metric: Additional Mortality 1-i1: TRUE means birds never directly impacted

-   Metric: Additional Mortality 1-i2: TRUE means birds directly impacted in some way at least once

-   Metric: Additional Mortality 1-i3: TRUE means birds displaced at least once, never barriered

-   Metric: Additional Mortality 1-i4: TRUE means birds barriered at least once, never displaced

-   Metric: Additional Mortality 1-i5: TRUE means birds barriered and displaced at least once

-   Metric: Additional Mortality 1-i6: All impacts grouped separately
